# MLM, Longitudinal: Hox ch 5 - student GPA


```{r, echo=FALSE}
library(pander)

panderOptions('digits', 2)
panderOptions('round', 2)
panderOptions('keep.trailing.zeros', TRUE)
panderOptions('keep.line.breaks', TRUE)
```


```{r}
library(tidyverse)    # all things tidy
library(pander)       # nice looking genderal tabulations
library(furniture)    # nice table1() descriptives
library(texreg)       # Convert Regression Output to LaTeX or HTML Tables
library(psych)        # contains some useful functions, like headTail
library(sjstats)      # ICC calculations
library(sjPlot)       # Visualization for Models
library(effects)      # Effect displays for Models

library(lme4)         # non-linear mixed-effects models

library(haven)        # read in SPSS dataset
```


## Background


```{block type='rmdlink', echo=TRUE}
The text **"Multilevel Analysis: Techniques and Applications, Third Edition"** [@hox2017] has a companion [website](https://multilevel-analysis.sites.uu.nl/) which includes links to all the data files used throughout the book (housed on the [book's GitHub repository](https://github.com/MultiLevelAnalysis)).  
```


The following example is used through out @hox2017's chapater 5.

The GPA for 200 college students were followed for 6 consecutive semesters (simulated).  Job status was also measured as number of hours worked for the same size occations.  Time-invariant covariates are the student's gender and high school GPA.  The variable `admitted` will not be used.  


```{r}
data_raw <- haven::read_sav("https://github.com/MultiLevelAnalysis/Datasets-third-edition-Multilevel-book/raw/master/chapter%205/GPA2/gpa2long.sav") %>% 
  haven::as_factor()             # retain the labels from SPSS --> factor

tibble::glimpse(data_raw) 
```

```{r}
data_raw %>% 
  dplyr::select(occas, job) %>% 
  table(useNA = "always")
```



```{r}
data_long <- data_raw %>% 
  dplyr::mutate(student = factor(student)) %>% 
  dplyr::mutate(sem = case_when(occas == "year 1 semester 1" ~ 1,
                                occas == "year 1 semester 2" ~ 2,
                                occas == "year 2 semester 1" ~ 3,
                                occas == "year 2 semester 2" ~ 4,
                                occas == "year 3 semester 1" ~ 5,
                                occas == "year 3 semester 2" ~ 6)) %>% 
  dplyr::mutate(semc = sem - 1) %>% 
  dplyr::mutate(job = fct_drop(job)) %>% 
  dplyr::mutate(hrs = case_when(job == "no job" ~ 0,
                                job == "1 hour" ~ 1,
                                job == "2 hours" ~ 2,
                                job == "3 hours" ~ 3,
                                job == "4 or more hours" ~ 4)) %>%  
  dplyr::select(student, sex, highgpa, sem, semc, job, hrs, gpa) %>% 
  dplyr::arrange(student, sem)

psych::headTail(data_long, top = 10)
```


```{r}
data_wide <- data_long %>% 
  furniture::wide(v.names = c("job", "hrs", "gpa"),
                  timevar = "sem",
                  id = "student")


psych::headTail(data_wide)
```


```{r}
data_wide %>% 
  furniture::table1(highgpa, gpa.1, job.1, hrs.1,
                    splitby = ~ sex,
                    output = "html",
                    test = TRUE)
```



```{r}
data_wide %>% 
  furniture::table1(gpa.1, gpa.2, gpa.3, gpa.4, gpa.5, gpa.6,
                    splitby = ~ sex,
                    output = "html")
```

## MLM

### Null Models and ICC



```{r}
fit_lmer_0_re <- lme4::lmer(gpa ~ 1 + (1|student),
                            data = data_long,
                            REML = TRUE)
```

```{r}
performance::icc(fit_lmer_0_re)
```


Over a third of the variance in the 6 GPA measures is variance between individuals, and about two-thirds is variance within individuals across time.


```{r}
fit_lmer_1_re <- lme4::lmer(gpa ~ semc + (1|student),
                            data = data_long,
                            REML = TRUE)
```

```{r}
performance::icc(fit_lmer_1_re)
```

After accounting for the linear change in GPA over semesters, about half of the remaining variance in GPA scores is attributable person-to-person differences.


### Fixed Effects


```{r}
fit_lmer_0_ml <- lme4::lmer(gpa ~ 1 + (1|student),
                            data = data_long,
                            REML = FALSE)

fit_lmer_1_ml <- lme4::lmer(gpa ~ semc + (1|student),
                            data = data_long,
                            REML = FALSE)


fit_lmer_2_ml <- lme4::lmer(gpa ~ semc + hrs + (1|student),
                            data = data_long,
                            REML = FALSE)


fit_lmer_3_ml <- lme4::lmer(gpa ~ semc + hrs + highgpa + sex + (1|student),
                            data = data_long,
                            REML = FALSE)
```


```{r}
screenreg(list(fit_lmer_0_ml, fit_lmer_1_ml, fit_lmer_2_ml, fit_lmer_3_ml),
          digits = 3)
```

```{r}
anova(fit_lmer_0_ml, fit_lmer_1_ml, fit_lmer_2_ml, fit_lmer_3_ml)
```

### Variance Explained by linear TIME at Level ONE

```{r}
lme4::VarCorr(fit_lmer_0_ml)  # baseilne
```

```{r}
lme4::VarCorr(fit_lmer_1_ml)  # model to compare
```


```{r}
lme4::VarCorr(fit_lmer_0_ml) %>%  # baseilne
  print(comp = c("Variance", "Std.Dev"),
        digits = 3)
```

```{r}
lme4::VarCorr(fit_lmer_0_ml) %>%  # model to compare
  print(comp = c("Variance", "Std.Dev"),
        digits = 3)
```



#### Raudenbush and Bryk

* Explained variance is a proportion of first-level variance only
* A good option when the multilevel sampling process is is close to two-stage simple random sampling

```{block type='genericEq', echo=TRUE}
**Raudenbush and Bryk Approximate Formula - Level 1 ** *approximate*
$$
approx \;R^2_1 = \frac{\sigma^2_{e-BL} - \sigma^2_{e-MC}}
             {\sigma^2_{e-BL} }
\tag{Hox 4.8}
$$
```



```{r}
(0.098 - 0.058) / 0.098
```


#### Snijders and Bosker


```{block type='genericEq', echo=TRUE}
**Snijders and Bosker Formula - Level 1 ** 

Random Intercepts Models Only, *address potential negative $R^2$ issue*
$$
R^2_1 = 1 - \frac{\sigma^2_{e-MC} + \sigma^2_{u0-MC}}
                 {\sigma^2_{e-BL} + \sigma^2_{u0-BL}}
$$
```




```{r}
1 - (0.058 + 0.063)/(0.098 + 0.057)
```


### Variance Explained by linear TIME at Level TWO

#### Raudenbush and Bryk

```{block type='genericEq', echo=TRUE}
**Raudenbush and Bryk Approximate Formula - Level 2 ** 
$$
approx \; R^2_s = \frac{\sigma^2_{u0-BL} - \sigma^2_{u0-MC}}
             {\sigma^2_{u0-BL} }
\tag{Hox 4.9}
$$
```





```{r}
(0.057 - 0.063)/ 0.057
```

YIKES!  Negative Variance explained!


#### Snijders and Bosker


```{block type='genericEq', echo=TRUE}
**Snijders and Bosker Formula Extended - Level 2 ** 
$$
R^2_2 = 1 - \frac{\frac{\sigma^2_{e-MC}}{B} + \sigma^2_{u0-MC}}
                 {\frac{\sigma^2_{e-BL}}{B} + \sigma^2_{u0-BL}}
$$

$B$ is the average size of the Level 2 units.  Technically, you should use the *harmonic mean*, but unless the clusters differ greatly in size, it doesn't make a huge difference.
```


```{r}
1 - (0.058/6 + 0.063) / (0.098/6 + 0.057)
```

Reason: The intercept only model overestimates the variance at the occasion level and underestimates the variance at the subject level (se chapter 4 of @hox2017)


### Random Effects



```{r}
fit_lmer_3_re <- lme4::lmer(gpa ~ semc + hrs + highgpa + sex + (1|student),
                            data = data_long,
                            REML = TRUE)


fit_lmer_4_re <- lme4::lmer(gpa ~ semc + hrs + highgpa + sex + (semc|student),
                            data = data_long,
                            REML = TRUE)
```


```{r}
anova(fit_lmer_3_re, fit_lmer_4_re, refit = FALSE)
```

### Cross-Level Interaction

```{r}
fit_lmer_4_ml <- lme4::lmer(gpa ~ semc + hrs + highgpa + sex + (semc|student),
                            data = data_long,
                            REML = FALSE)

fit_lmer_4_re <- lme4::lmer(gpa ~ semc + hrs + highgpa + sex + (semc|student),
                            data = data_long,
                            REML = TRUE)



fit_lmer_5_ml <- lme4::lmer(gpa ~ semc + hrs + highgpa + sex + semc:sex + (semc|student),
                            data = data_long,
                            REML = FALSE)

fit_lmer_5_re <- lme4::lmer(gpa ~ semc + hrs + highgpa + sex + semc:sex + (semc|student),
                            data = data_long,
                            REML = FALSE)
```


```{r}
texreg::screenreg(list(fit_lmer_4_ml, fit_lmer_5_ml, fit_lmer_5_re))
```


```{r}
anova(fit_lmer_4_ml, fit_lmer_5_ml)
```


```{r}
sjPlot::plot_model(fit_lmer_5_re,
                   type = "pred",
                   terms = c("semc", "sex"))
```





