# GEE, Continuous Outcome: Beat the Blues

```{r, include=FALSE}
knitr::opts_chunk$set(comment     = "",
                      echo        = TRUE, 
                      warning     = FALSE, 
                      message     = FALSE,
                      fig.align   = "center", # center all figures
                      fig.width   = 6,        # set default figure width to 4 inches
                      fig.height  = 4)        # set default figure height to 3 inches
```

```{r, message=FALSE, error=FALSE}
library(tidyverse)    # all things tidy
library(pander)       # nice looking genderal tabulations
library(furniture)    # nice table1() descriptives
library(texreg)       # Convert Regression Output to LaTeX or HTML Tables
library(psych)        # contains some useful functions, like headTail

library(lme4)         # Linear, generalized linear, & nonlinear mixed models

library(corrplot)     # Vizualize correlation matrix
library(gee)          # Genderalized Estimation Equation Solver
library(geepack)      # Genderalized Estimation Equation Package 
library(MuMIn)        # Multi-Model Inference (caluclate QIC)

library(HSAUR)        # package with the dataset
```

## Prepare and get to know the dataset

### Read in the data

This data set is in the **HSAUR** package.  It is from a clinical trial of an interactive multimedia program called 'Beat the Blues'. The variables are as follows:   

  * `drug` did the patient take anti-depressant drugs (No or Yes)    
  * `length`  the length of the current episode of depression, a factor with levels:    
    + "<6m" *less than six months*     
    + ">6m" *more than six months*   
  * `treatment` treatment group, a factor with levels:   
    + "TAU" *treatment as usual*     
    + "BtheB" *Beat the Blues*    
  * `bdi.pre` Beck Depression Inventory II, before treatment    
  * `bdi.2m` Beck Depression Inventory II, after 2 months    
  * `bdi.4m` Beck Depression Inventory II, after 4 months    
  * `bdi.6m` Beck Depression Inventory II, after 6 months    
  * `bdi.8m` Beck Depression Inventory II, after 8 months    


```{r}
data(BtheB, package = "HSAUR")

BtheB %>% 
  psych::headTail()
```


### Tidy up the dataset

```{r}
btb_wide <- BtheB %>% 
  dplyr::mutate(id = row_number()) %>%           # create a new variable to ID participants
  dplyr::select(id, treatment,                   # specify that ID variable is first
                bdi.pre, bdi.2m, bdi.4m, bdi.6m, bdi.8m)
```


```{r, results='asis'}
btb_wide %>% 
  furniture::table1(bdi.pre, bdi.2m, bdi.4m, bdi.6m, bdi.8m,
                    splitby = ~ treatment,
                    test = TRUE,
                    output = "html")
```

### Restructure to long format

```{r}
btb_long <- btb_wide %>% 
  tidyr::gather(key   = month,
                value = bdi,
                starts_with("bdi")) %>% 
  dplyr::mutate(month = case_when(month == "bdi.pre" ~ 0,
                                  month == "bdi.2m"  ~ 2,
                                  month == "bdi.4m"  ~ 4,
                                  month == "bdi.6m"  ~ 6,
                                  month == "bdi.8m"  ~ 8)) %>% 
  dplyr::filter(complete.cases(id, bdi, treatment, month)) %>% 
  dplyr::arrange(id, month) %>% 
  dplyr::select(id, treatment, month, bdi)
```


```{r}
btb_long %>% 
  psych::headTail(top = 10, bottom = 10) 
```


### Visualize: Person-profile Plots

Create spaghetti plots of the raw, observed data

```{r}
btb_long %>% 
  ggplot(aes(x = month,
             y = bdi)) +
  geom_point() +
  geom_line(aes(group = id), 
            size = 1, 
            alpha = 0.3) +
  geom_smooth(method = "lm") +
  facet_grid(.~ treatment) +
  theme_bw() +
  labs(title = "BtheB - Observed Data Across Time with LM Smoother")
```



```{r}
btb_long %>% 
  ggplot(aes(x = month,
             y = bdi,
             color = treatment,
             fill = treatment)) +
  geom_smooth(method = "lm") +
  theme_bw() +
  labs(title = "BtheB - Observed Data: Linear Smoothers Across Time")
```

### Calculate the Observed Correlation Structure


```{r}
bdi_corr <- btb_wide %>% 
  dplyr::select(starts_with("bdi")) %>% 
  stats::cor(use="pairwise.complete.obs")

bdi_corr
```


### Plot the correlation matrix to get a better feel for the pattern

```{r}
corrplot::corrplot.mixed(bdi_corr)
```




## Multiple Regression (OLS)

This ignores any correlation between repeated measures on the same individual and treats all observations as independent.

### Fit the model and view the output in a table

```{r}
btb_lm <- stats::lm(bdi ~ treatment + month,
                    data = btb_long)
```


```{r, include=FALSE}
texreg::screenreg(btb_lm)
```

### Parameter Estimates

```{r, results='asis'}
# Knit to Website: texreg::htmlreg()
# Knit to PDF:     texreg::texreg()
# View on Screen:  texreg::screenreg()

texreg::htmlreg(btb_lm,
                custom.model.names = "OLS")
```

### Plot the model predictions

```{r}
effects::Effect(focal.predictors = c("treatment", "month"), 
                mod = btb_lm) %>% 
  data.frame %>% 
  dplyr::mutate(treatment = fct_reorder2(treatment, month, fit)) %>%  
  ggplot(aes(x = month,
             y = fit,
             color = treatment,
             fill = treatment)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower,
                  ymax = upper), alpha = 0.3) +
  theme_bw() +
  labs(title = "BtheB - Predictions from a Linear Model (lm)") +
  theme(legend.position = c(1, 1),
        legend.justification = c(1.1, 1.1),
        legend.background = element_rect(color = "black"))
```


## Multilevel Models (MLM)

### Fit the model and compare to the prior model


```{r}
btb_lmer_RI   <- lme4::lmer(bdi ~ treatment + month + (1 | id), 
                            data = btb_long, 
                            REML = TRUE)

btb_lmer_RIAS <- lme4::lmer(bdi ~ treatment + month + (month | id), 
                            data = btb_long, 
                            REML = TRUE)
```

```{r, include=FALSE}
texreg::screenreg(list(btb_lm, btb_lmer_RI, btb_lmer_RIAS), 
          custom.model.names = c("OLS", "MLM-RI", "MLM-RIAS"))
```

```{r, results='asis'}
# Knit to Website: texreg::htmlreg()
# Knit to PDF:     texreg::texreg()
# View on Screen:  texreg::screenreg()

texreg::htmlreg(list(btb_lm, btb_lmer_RI, btb_lmer_RIAS), 
                custom.model.names = c("OLS", "MLM-RI", "MLM-RIAS"))
```

### Likelihood Ratio Test


```{r}
anova(btb_lmer_RI, 
      btb_lmer_RIAS, 
      refit = FALSE)
```


### Plot the model predictions

```{r}
effects::Effect(c("treatment", "month"), mod = btb_lmer_RI) %>% 
  data.frame %>%   
  dplyr::mutate(treatment = fct_reorder2(treatment, month, fit)) %>%
  ggplot(aes(x = month,
             y = fit,
             color = treatment,
             fill = treatment)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower,
                  ymax = upper), alpha = 0.3) +
  theme_bw() +
  labs(title = "BtheB - Predictions from a Multilevel Model (lmer)") +
  theme(legend.justification = c(0, 0),
        legend.position = c(0, 0),
        legend.background = element_rect(color = "black"))
```




## General Estimating Equations, GEE


### Fit the models and compare to the prior model

Use the `gee()` function from the **gee** package for the results to be used in a `texreg()` table.

```{r}
btb_gee_in <- gee::gee(bdi ~ treatment + month, 
                       data = btb_long, 
                       id = id, 
                       family = gaussian, 
                       corstr = 'independence')

btb_gee_ex <- gee::gee(bdi ~ treatment + month, 
                       data = btb_long, 
                       id = id, 
                       family = gaussian, 
                       corstr = 'exchangeable')

# The AR-1 fails if any subjects have only 1 observation
# to use this one, we would need to remove participants with only 1 BDI
# btb_gee_ar <- gee(bdi ~ treatment + month, 
#                  data = btb_long, 
#                  id = id, 
#                  family = gaussian, 
#                  corstr = 'AR-M',
#                  Mv = 1)

btb_gee_un <- gee::gee(bdi ~ treatment + month, 
                       data = btb_long, 
                       id = id, 
                       family = gaussian, 
                       corstr = 'unstructured')
```





### Parameter Estimates

```{r, include=FALSE}
texreg::screenreg(list(btb_lm, 
                      btb_lmer_RI, 
                      btb_gee_in, 
                      btb_gee_ex, 
                      btb_gee_un))
```


```{r, results='asis'}
# Knit to Website: texreg::htmlreg()
# Knit to PDF:     texreg::texreg()
# View on Screen:  texreg::screenreg()

texreg::htmlreg(list(btb_lm, 
                      btb_lmer_RI, 
                      btb_gee_in, 
                      btb_gee_ex, 
                      btb_gee_un),
                  custom.model.names = c("OLS", 
                                        "MLM-RI", 
                                        "GEE-in", 
                                        "GEE-ex", 
                                        "GEE-un"))
```





#### Re-Fit Models

Use the `geeglm()` function from the **geepack** package for the results to be used in a `anova()` table.


```{r}
btb_geeglm_in <- geeglm(bdi ~ treatment + month, 
                        data = btb_long, 
                        id = id,
                        wave = month,
                        family = gaussian, 
                        corstr = 'independence')

btb_geeglm_ex <- geeglm(bdi ~ treatment + month, 
                        data = btb_long, 
                        id = id, 
                        wave = month,
                        family = gaussian, 
                        corstr = 'exchangeable')


btb_geeglm_ar <- geeglm(bdi ~ treatment + month, 
                        data = btb_long, 
                        id = id, 
                        wave = month,
                        family = gaussian, 
                        corstr = 'ar1')

btb_geeglm_un <- geeglm(bdi ~ treatment + month, 
                        data = btb_long, 
                        id = id,
                        wave = month,
                        family = gaussian, 
                        corstr = 'unstructured')
```


### Can't Use the Likelihood Ratio Test

The `anova()` function is used to compare nested models for parameters (fixed effects), not correlation structures.

```{r}
anova(btb_geeglm_in, btb_geeglm_ex)
anova(btb_geeglm_in, btb_geeglm_ar)
anova(btb_geeglm_in, btb_geeglm_un)
```



### Variaous QIC Measures of Fit

References:

* Pan, W. 2001. Akaike's information criterion in generalized estimating equations. Biometrics 57:120-125. https://onlinelibrary.wiley.com/doi/abs/10.1111/j.0006-341X.2001.00120.x 

* Burnham, K. P. and D. R. Anderson. 2002. Model selection and multimodel inference: a practical information-theoretic approach. Second edition. Springer Science and Business Media, Inc., New York.
https://cds.cern.ch/record/1608735/files/9780387953649_TOC.pdf 



The `QIC()` is one way to **try** to measure model fit.  You can enter more than one model into a single function call.

QIC(I) based on independence model  <-- suggested by Pan (Biometric, March 2001), asymptotically unbiased estimator (choose the correlation stucture that produces the smallest QIC(I), p122)

```{r}
MuMIn::QIC(btb_geeglm_in, 
           btb_geeglm_ex, 
           btb_geeglm_ar, 
           btb_geeglm_un, 
           typeR = FALSE) # default
```


QIC(R) is based on quasi-likelihood of a working correlation R model, can **NOT** be used to select the working correlation matrix.

```{r}
MuMIn::QIC(btb_geeglm_in, 
           btb_geeglm_ex, 
           btb_geeglm_ar, 
           btb_geeglm_un, 
           typeR = TRUE)    # NOT the default
```


QIC_U(R) approximates QIC(R), and while both are useful for variable selection, they can NOT be applied to select the working correlation matrix.

```{r}
MuMIn::QICu(btb_geeglm_in, 
            btb_geeglm_ex, 
            btb_geeglm_ar, 
            btb_geeglm_un) 
```





```{r}
MuMIn::model.sel(btb_geeglm_in, 
                 btb_geeglm_ex, 
                 btb_geeglm_ar, 
                 btb_geeglm_un, 
                 rank = "QIC")    #sorts the best to the TOP, uses QIC(I)
```






### Plot the model predictions

Do not worry about confidence intervals.

```{r}
expand.grid(treatment = levels(btb_long$treatment),
            month = seq(from = 0, to = 8, by = 2)) %>%  
  mutate(fit_in = predict(btb_geeglm_in,
                       newdata = .,
                       type = "response")) %>% 
  mutate(fit_ex = predict(btb_geeglm_ex,
                       newdata = .,
                       type = "response")) %>% 
  mutate(fit_ar = predict(btb_geeglm_ar,
                       newdata = .,
                       type = "response")) %>% 
  mutate(fit_un = predict(btb_geeglm_un,
                       newdata = .,
                       type = "response")) %>% 
  gather(key = variable,
         value = fit,
         starts_with("fit")) %>% 
  separate(col = variable,
           into = c("junk", "covR")) %>% 
  select(-junk) %>% 
  mutate(covR = factor(covR, 
                       levels = c("un", "ar", "ex", "in"),
                       labels = c("Unstructured",
                                  "Auto-Regressive",
                                  "Compound Symetry",
                                  "Independence"))) %>% 
  ggplot(aes(x = month,
             y = fit,
             linetype = treatment)) +
  geom_line(alpha = 0.6) +
  theme_bw() +
  labs(title    = "BtheB - Predictions from four GEE models (geeglm)",
       linetype = "Treatment",
       color    = "Corelation\nStructure",
       size     = "Corelation\nStructure") +
  scale_linetype_manual(values = c("solid", "longdash")) +
  scale_size_manual(values     = c(2, 1, 1, 1)) +
  scale_color_manual(values    = c("red", 
                                   "dodgerblue",
                                   "blue",
                                   "darkgreen")) +
  theme(legend.key.width = unit(1, "cm")) +
  facet_wrap(~ covR)
```








