# GEE, Binary Outcome: Respiratory Illness

```{r, include=FALSE}
knitr::opts_chunk$set(comment     = "",
                      echo        = TRUE, 
                      warning     = FALSE, 
                      message     = FALSE,
                      fig.align   = "center", # center all figures
                      fig.width   = 6,        # set default figure width to 4 inches
                      fig.height  = 4)        # set default figure height to 3 inches
```

```{r, message=FALSE, error=FALSE}
library(tidyverse)    # all things tidy
library(pander)       # nice looking genderal tabulations
library(furniture)    # nice table1() descriptives
library(texreg)       # Convert Regression Output to LaTeX or HTML Tables
library(psych)        # contains some useful functions, like headTail

library(lme4)         # Linear, generalized linear, & nonlinear mixed models

library(corrplot)     # Vizualize correlation matrix
library(gee)          # Genderalized Estimation Equation Solver
library(geepack)      # Genderalized Estimation Equation Package 
library(MuMIn)        # Multi-Model Inference (caluclate QIC)

library(HSAUR)        # package with the dataset
```

## Prepare and get to know the dataset

### Read in the data

This data set is in the **HSAUR** package. 

```{r}
data("respiratory", package = "HSAUR")

str(respiratory)

psych::headTail(respiratory)
```

### Wide Format

```{r}
data_wide <- respiratory %>% 
  tidyr::spread(key = month,
                value = status,
                sep = "_") %>% 
  dplyr::rename("BL_status" = "month_0") %>% 
  dplyr::arrange(subject) %>% 
  dplyr::select(subject, centre, sex, age, treatment, BL_status, starts_with("month"))  

str(data_wide)

psych::headTail(data_wide)
```

### Long Format

```{r}
data_long <- data_wide%>% 
  tidyr::gather(key = month,
                value = status,
                starts_with("month")) %>% 
  dplyr::mutate(month = str_sub(month, start = -1) %>% as.numeric) %>% 
  dplyr::mutate(status = case_when(status == "poor" ~ 0,
                                   status == "good" ~ 1)) %>% 
  dplyr::arrange(subject, month) %>% 
  dplyr::select(subject, centre, sex, age, treatment, BL_status, month, status) 


str(data_long)

psych::headTail(data_long)
```


## Exploratory Data Analysis

### Summary Statustics

#### Demographics and Baseline Measure


```{r}
data_wide %>% 
  furniture::table1(centre, sex, age, BL_status, 
                    splitby = ~ treatment,
                    output = "html")
```

#### Status  Over Time

```{r}
data_wide %>% 
  furniture::table1(month_1, month_2, month_3, month_4, 
                    splitby = ~ treatment,
                    output = "html")
```


```{r}
data_month_trt_prop <- data_long %>% 
  dplyr::group_by(treatment, month) %>% 
  dplyr::summarise(n = n(),
                   prop_good = mean(status),
                   prop_sd = sd(status),
                   prop_se = prop_sd/n)

psych::headTail(data_month_trt_prop)
```


### Visualization

#### Status Over Time


```{r}
data_month_trt_prop %>% 
  ggplot(aes(x = month,
             y = prop_good,
             group = treatment,
             color = treatment)) +
  geom_errorbar(aes(ymin = prop_good - prop_se,
                    ymax = prop_good + prop_se),
                width = .25) +
  geom_point() +
  geom_line() +
  theme_bw()
```

## Logisitc Regression


```{r}
resp_glm <- glm(status ~ centre + treatment + sex + BL_status + I(age-33) + I((age-33)^2),
                data = data_long,
                family = binomial(link = "logit"))

summary(resp_glm)
```


```{r}
sjPlot::tab_model(resp_glm)
```


## GEE

### Indepdendence

```{r}
resp_gee_in <- gee::gee(status ~ centre + treatment + sex + BL_status + I(age-33) + I((age-33)^2),
                        data = data_long,
                        family = binomial(link = "logit"),
                        id = subject,
                        corstr = "independence",
                        scale.fix = TRUE,
                        scale.value = 1)

summary(resp_gee_in)
```

The results for GEE fit with the independence correlation structure produces results that are nearly identical to the GLM model.

The robust (sandwhich) standard errors are however much larger than the naive stadard errors


### Exchangeable

```{r}
resp_gee_ex <- gee::gee(status ~ centre + treatment + sex + BL_status + I(age-33) + I((age-33)^2),
                        data = data_long,
                        family = binomial(link = "logit"),
                        id = subject,
                        corstr = "exchangeable",
                        scale.fix = TRUE,
                        scale.value = 1)

summary(resp_gee_ex)
```

Notice that the naive stadard errors are more similar to the robust (sandwhich) standard errors, supporting that this is a better fitting model

### Paramgeter Estimates Table

The GEE models display the robus (sandwhich) standard errors.


```{r}
texreg::screenreg(list(resp_glm, resp_gee_in, resp_gee_ex),
                  custom.model.names = c("GLM", "GEE-INDEP", "GEE-EXCH"),
                  caption = "Estimates = log odds ratios",
                  digits = 4)
```

Comparing the two GEE models: parameters are identical and so are the robust (sandwhich) standard errors.



```{r}
summary(resp_gee_ex) %>% names
```

```{r}
summary(resp_gee_ex)$coeff
```


```{r}
resp_gee_ex %>% coef() %>% exp()
```



```{r}
summary(resp_gee_ex)$coeff["treatmenttreatment", "Estimate"]
```

```{r}
OR_trt_EST <- summary(resp_gee_ex)$coeff["treatmenttreatment", "Estimate"] %>% exp()
OR_trt_EST
```


```{r}
OR_Trt_SE <- summary(resp_gee_ex)$coeff["treatmenttreatment", "Robust S.E."] %>% exp
OR_Trt_SE
```

```{r}
OR_trt_EST + c(-1, +1)*1.96*OR_Trt_SE
```






### Refit with the `geepack` package


```{r}
resp_geeglm_ex <- geepack::geeglm(status ~ centre + treatment + sex + BL_status + I(age-33) + I((age-33)^2),
                        data = data_long,
                        family = binomial(link = "logit"),
                        id = subject,
                        waves = month,
                        corstr = "exchangeable")
```


```{r}
summary(resp_geeglm_ex)
```

```{r}
resp_geeglm_ex %>% coef() %>% exp()
```







### Visualize the Model

```{r}
summary(data_long)
```

```{r}
expand.grid(centre = "1",
            treatment = levels(data_long$treatment),
            sex = "female",
            age = seq(from = 11, to = 68, by = 1),
            BL_status = levels(data_long$BL_status)) %>% 
  dplyr::mutate(fit = predict(resp_geeglm_ex,
                              newdata = .,
                              type = "response")) %>% 
  ggplot(aes(x = age,
             y = fit,
             linetype = treatment)) +
  geom_line() +
  theme_bw() + 
  facet_grid(.~ BL_status) +
  labs(x = "Age, years",
       y = "Predicted Probability of GOOD Respiratory Status",
       title = "For Females at Center 1")
```


```{r}
expand.grid(centre = "2",
            treatment = levels(data_long$treatment),
            sex = "male",
            age = seq(from = 11, to = 68, by = 1),
            BL_status = levels(data_long$BL_status)) %>% 
  dplyr::mutate(fit = predict(resp_geeglm_ex,
                              newdata = .,
                              type = "response")) %>% 
  ggplot(aes(x = age,
             y = fit,
             linetype = treatment)) +
  geom_line() +
  theme_bw() + 
  facet_grid(.~ BL_status) +
  labs(x = "Age, years",
       y = "Predicted Probability of GOOD Respiratory Status",
       title = "For Males at Center 2")
```

