# GEE, Count Outcome: Epilepsy

```{r, include=FALSE}
knitr::opts_chunk$set(comment     = "",
                      echo        = TRUE, 
                      warning     = FALSE, 
                      message     = FALSE,
                      fig.align   = "center", # center all figures
                      fig.width   = 6,        # set default figure width to 4 inches
                      fig.height  = 4)        # set default figure height to 3 inches
```

## Packages

### CRAN

```{r, message=FALSE, error=FALSE}
library(tidyverse)    # all things tidy
library(pander)       # nice looking genderal tabulations
library(furniture)    # nice table1() descriptives
library(texreg)       # Convert Regression Output to LaTeX or HTML Tables
library(psych)        # contains some useful functions, like headTail

library(lme4)         # Linear, generalized linear, & nonlinear mixed models

library(corrplot)     # Vizualize correlation matrix
library(gee)          # Genderalized Estimation Equation Solver
library(geepack)      # Genderalized Estimation Equation Package 
library(MuMIn)        # Multi-Model Inference (caluclate QIC)

library(HSAUR)        # package with the dataset
```


### GitHub

Helper `extract` functions for exponentiating parameters form generalized regression models within a `texreg` table of model parameters.

```{r, message=FALSE, error=FALSE}
# install.packages("devtools")
# library(devtools)
# install_github("SarBearSchwartz/texreghelpr")

library(texreghelpr)
```

## Prepare and get to know the dataset


> This data set is in the **HSAUR** package. 

In this clinical trial, 59 patients suffering from epilepsy were randomized to groups receiving either the anti-epileptic drug "Progabide"" or a placebo in addition to standard chemotherapy. The numbers of seizures suffered in each of four, two-week periods were recorded for each patient along with a baseline seizure count for the 8 weeks prior to being randomized to treatment and age. The main question of interest is whether taking progabide reduced the number of epileptic seizures compared with placebo.

* Indicators    

    + `subject` the patient ID, a factor with levels 1 to 59    
    + `period` treatment period, an ordered factor with levels 1 to 4
    
    
* Outcome or dependent variable    

    +`seizure.rate` the number of seizures (2-weeks)

* Main predictor or independent variable of interest   

    + `treatment` the treatment group, a factor with levels placebo and Progabide


* Time-invariant Covariates    

    + `age` the age of the patient    
    + `base` the number of seizures before the trial (8 weeks)


### Read in the data

```{r}
data("epilepsy", package = "HSAUR")
```


Problem:  The outcome (`seizure.rate`) were counts over a TWO-week period and we would like to interpret the results in terms of effects on the WEEKLY rate.

* If we divide the values by TWO to get weekly rates, the outcome might be a decimal number
* The Poisson distribution may only be used for whole numbers (not deciamls)

Solution:  We need to include an `offset` term in the model that indicates the LOG DURATION of each period.

* Every observation period is exactly 2 weeks in this experiment
* Create a variable in the original dataset that is equal to the  LOG DURATION (`per = log(2)`)
* To the formula for the `glm()` or `gee()`, add: ` + offset(per)`



### Long Format

```{r}
data_long <- epilepsy %>% 
  dplyr::select(subject, age, treatment, base, 
                period, seizure.rate) %>% 
  dplyr::mutate(per = log(2)) %>%                  # new variable to use with the offset
  dplyr::mutate(base_wk = base/8)
  
str(data_long)

psych::headTail(data_long, top = 10, bottom = 6)
```



### Wide Format

```{r}
data_wide <- data_long %>% 
  tidyr::spread(key = period,
                value = seizure.rate,
                sep = "_") %>% 
  dplyr::arrange(subject)

str(data_wide)

psych::headTail(data_wide)
```

## Exploratory Data Analysis

### Summarize

#### Demographics and Baseline

```{r}
data_wide %>% 
  furniture::table1(age, base, base_wk,
                    splitby = ~ treatment,
                    test = TRUE,
                    type = "full",
                    output = "markdown")
```

#### Outcome Across Time

> Note: The Poisson distribution specifies that the **MEAN = VARIANCE**

In this dataset, the variance is much larger than the mean, at all time points for both groups.  This is evidence of **overdispersion** and suggest the scale parameter should be **greater than one**.

```{r}
data_long %>% 
  dplyr::group_by(treatment, period) %>% 
  dplyr::summarise(N = n(),
                   M = mean(seizure.rate),
                   VAR = var(seizure.rate),
                   SD = sd(seizure.rate)) %>% 
  pander::pander()
```

#### Correlation Across Time

Raw Scale

```{r}
data_long %>% 
  dplyr::select(subject, period, seizure.rate ) %>% 
  tidyr::spread(key = period,
                value = seizure.rate ) %>% 
  dplyr::select(-subject) %>% 
  cor() %>% 
  corrplot::corrplot.mixed()
```

Log Scale

```{r}
data_long %>% 
  dplyr::mutate(rate_wk = log(seizure.rate + 1)) %>% 
  dplyr::select(subject, period, rate_wk) %>% 
  tidyr::spread(key = period,
                value = rate_wk) %>% 
  dplyr::select(-subject) %>% 
  cor() %>% 
  corrplot::corrplot.mixed()
```



### Visualize

#### Oucome on the Raw Scale

There appear to be quite a few extreme values or outliers, particularly for the Progabide group during period one.

Since the outcome is truely a **COUNT**, we will model it with a **Poisson** distribution combined with a **LOG** link.

```{r}
data_long %>% 
  ggplot(aes(x = period,
             y = seizure.rate)) +
  geom_boxplot() +
  theme_bw() +
  facet_grid(.~ treatment)
```

To investigate possible outliers, we should transform the outcome with the `log` function first.  

> Note: Since some participants reported no seizures during a two week period and the `log(0)` is **unndefinded**, we must add some amount to the values before transforming.  Here we have chosen to add the value of $1$.

```{r}
data_long %>% 
  ggplot(aes(x = period,
             y = log(seizure.rate + 1))) +
  geom_boxplot() +
  
  theme_bw() +
  facet_grid(.~ treatment)
```

```{r}
data_long %>% 
  ggplot(aes(x = period,
             y = log(seizure.rate + 1))) +
  geom_line(aes(group = subject)) +
  theme_bw() +
  facet_grid(.~ treatment)
```


```{r}
data_long %>% 
  ggplot(aes(x = period,
             y = log(seizure.rate + 1))) +
  geom_smooth(aes(group = subject),
              method = "lm",
              se = FALSE) +
  geom_smooth(aes(group = 1),
              color = "red",
              size = 1.5,
              method = "lm",
              se = FALSE) +
  theme_bw() +
  facet_grid(.~ treatment)
```


## Poisson Regression (GLM)


```{r}
fit_glm <- glm(seizure.rate ~ base + age + treatment + offset(per),
               data = data_long,
               family = poisson(link = "log"))

summary(fit_glm)
```

## Generalized Estimating Equations (GEE)


### Match Poisson Regresssion (GLM)

* correlation structure: `independence`
* scale parameter = $1$

```{r}
fit_gee_ind_s1 <- gee::gee(seizure.rate ~ base + age + treatment + offset(per),
                           data = data_long,
                           family = poisson(link = "log"),
                           id = subject,
                           corstr = "independence",
                           scale.fix = TRUE,
                           scale.value = 1)

summary(fit_gee_ind_s1)
```

* The estimates and the naive standard errors match the GLM exactly.  

* The naive SE's are much smaller (*half*) than the robust (sandwich) SE's, suggesting a poor fit.


### Change Correlation Sturucture

* correlation structure: `exchangeable`
* scale parameter = $1$

```{r}
fit_gee_exc_s1 <- gee::gee(seizure.rate ~ base + age + treatment + offset(per),
                           data = data_long,
                           family = poisson(link = "log"),
                           id = subject,
                           corstr = "exchangeable",
                           scale.fix = TRUE,
                           scale.value = 1)

summary(fit_gee_exc_s1)
```

* Although the estimated beta parameters are not much different, the naive SE's are some closer to the robust SE's.

### Estimate the Additional Scale Parameter

* correlation structure: `exchangeable`
* scale parameter = *freely estimated*

```{r}
fit_gee_exc_sf <- gee::gee(seizure.rate ~ base + age + treatment + offset(per),
                           data = data_long,
                           family = poisson(link = "log"),
                           id = subject,
                           corstr = "exchangeable",
                           scale.fix = FALSE)

summary(fit_gee_exc_sf)
```

* The naive SE's are much closer inline with the robust SE's.

* The sclae parameter is estimated to be much larger than $1$.


### Compare Models


#### Raw Estimates (logit scale) 

```{r, include=FALSE}
texreg::screenreg(list(fit_glm, fit_gee_ind_s1, fit_gee_exc_s1, fit_gee_exc_sf),
                  digits = 3,
                  caption = "Estimates on Logit Scale")
```



```{r, results="asis"}
# Knit to Website: texreg::htmlreg()
# Knit to PDF:     texreg::texreg()
# View on Screen:  texreg::screenreg()

texreg::htmlreg(list(fit_glm, 
                     fit_gee_ind_s1, 
                     fit_gee_exc_s1, 
                     fit_gee_exc_sf),
                custom.model.names = c("GLM",
                                       "GEE-Indep(1)",
                                       "GEE-Exchg(1)",
                                       "GEE-Exchg(free)"),
                caption = "Estimates on Logit Scale",
                digits = 3)
```



#### Exponentiate the Estimates (odds ratio scale)


```{r, include=FALSE}
texreg::screenreg(list(extract_glm_exp(fit_glm), 
                       extract_gee_exp(fit_gee_ind_s1), 
                       extract_gee_exp(fit_gee_exc_s1), 
                       extract_gee_exp(fit_gee_exc_sf)),
                custom.model.names = c("GLM",
                                       "GEE-Indep(1)",
                                       "GEE-Exchg(1)",
                                       "GEE-Exchg(free)"),
                  digits = 3,
                  caption = "Estimates on Odds-Ratio Scale",
                  caption.above = TRUE,
                  ci.test = 1)
```


```{r, results="asis"}
# Knit to Website: texreg::htmlreg()
# Knit to PDF:     texreg::texreg()
# View on Screen:  texreg::screenreg()

texreg::htmlreg(list(extract_glm_exp(fit_glm), 
                     extract_gee_exp(fit_gee_ind_s1), 
                     extract_gee_exp(fit_gee_exc_s1), 
                     extract_gee_exp(fit_gee_exc_sf)),
                custom.model.names = c("GLM",
                                       "GEE-Indep(1)",
                                       "GEE-Exchg(1)",
                                       "GEE-Exchg(free)"),
                digits = 3,
                caption = "Estimates on Odds-Ratio Scale",
                caption.above = TRUE,
                ci.test = 1)
```

> CONCLUSION:
>
> There is no evidence that Progabide effects the weekly rate of epileptic seizures differently than placebo.










