# GEE, Count Outcome: Epilepsy

```{r, include=FALSE}
knitr::opts_chunk$set(comment     = "",
                      echo        = TRUE, 
                      warning     = FALSE, 
                      message     = FALSE,
                      fig.align   = "center", # center all figures
                      fig.width   = 6,        # set default figure width to 4 inches
                      fig.height  = 4)        # set default figure height to 3 inches
```

## Packages

### CRAN

```{r, message=FALSE, error=FALSE}
library(tidyverse)    # all things tidy
library(pander)       # nice looking genderal tabulations
library(furniture)    # nice table1() descriptives
library(texreg)       # Convert Regression Output to LaTeX or HTML Tables
library(psych)        # contains some useful functions, like headTail

library(lme4)         # Linear, generalized linear, & nonlinear mixed models

library(corrplot)     # Vizualize correlation matrix
library(gee)          # Genderalized Estimation Equation Solver
library(geepack)      # Genderalized Estimation Equation Package 
library(MuMIn)        # Multi-Model Inference (caluclate QIC)

library(HSAUR)        # package with the dataset
```


### GitHub

Helper `extract` functions for exponentiating parameters form generalized regression models within a `texreg` table of model parameters.

```{r, message=FALSE, error=FALSE}
# remotes::install_github("sarbearschwartz/texreghelpr") # first time

library(texreghelpr)
```

## Background


> This dataset was used as an example in Chapter 11 of "A Handbook of Statistical Analysis using R" by Brian S. Everitt and Torsten Hothorn.  The authors include this data set in their **HSAUR** package on `CRAN`. 


**The Background**

In this clinical trial, 59 patients suffering from epilepsy were randomized to groups receiving either the anti-epileptic drug "Progabide"" or a placebo in addition to standard chemotherapy. The numbers of seizures suffered in each of four, two-week periods were recorded for each patient along with a baseline seizure count for the 8 weeks prior to being randomized to treatment and age. 


**The Research Question**

The main question of interest is whether taking progabide reduced the number of epileptic seizures compared with placebo.


**The Data**

* Indicators    

    + `subject` the patient ID, a factor with levels 1 to 59    
    + `period` treatment period, an ordered factor with levels 1 to 4
    
    
* Outcome or dependent variable    

    +`seizure.rate` the number of seizures (2-weeks)

* Main predictor or independent variable of interest   

    + `treatment` the treatment group, a factor with levels placebo and Progabide


* Time-invariant Covariates    

    + `age` the age of the patient    
    + `base` the number of seizures before the trial (8 weeks)


### Read in the data

```{r}
data("epilepsy", package = "HSAUR")
```


Problem:  The outcome (`seizure.rate`) were counts over a TWO-week period and we would like to interpret the results in terms of effects on the WEEKLY rate.

* If we divide the values by TWO to get weekly rates, the outcome might be a decimal number
* The Poisson distribution may only be used for whole numbers (not deciamls)

Solution:  We need to include an `offset` term in the model that indicates the LOG DURATION of each period.

* Every observation period is exactly 2 weeks in this experiment
* Create a variable in the original dataset that is equal to the  LOG DURATION (`per = log(2)`)
* To the formula for the `glm()` or `gee()`, add: ` + offset(per)`



### Long Format

```{r}
data_long <- epilepsy %>% 
  dplyr::select(subject, age, treatment, base, 
                period, seizure.rate) %>% 
  dplyr::mutate(per = log(2)) %>%                  # new variable to use with the offset
  dplyr::mutate(base_wk = base/8)
  
str(data_long)

psych::headTail(data_long, top = 10, bottom = 6)
```



### Wide Format

```{r}
data_wide <- data_long %>% 
  tidyr::spread(key = period,
                value = seizure.rate,
                sep = "_") %>% 
  dplyr::arrange(subject)

str(data_wide)

psych::headTail(data_wide)
```

## Exploratory Data Analysis

### Summarize

#### Demographics and Baseline

```{r}
data_wide %>% 
  dplyr::group_by(treatment) %>% 
  furniture::table1(age, base, base_wk,
                    total = TRUE,
                    test = TRUE,
                    digits = 2,
                    type = "full",
                    output = "markdown")
```

#### Outcome Across Time

> Note: The Poisson distribution specifies that the **MEAN = VARIANCE**

In this dataset, the variance is much larger than the mean, at all time points for both groups.  This is evidence of **overdispersion** and suggest the scale parameter should be **greater than one**.

```{r}
data_long %>% 
  dplyr::group_by(treatment, period) %>% 
  dplyr::summarise(N = n(),
                   M = mean(seizure.rate),
                   VAR = var(seizure.rate),
                   SD = sd(seizure.rate)) %>% 
  pander::pander()
```

#### Correlation Across Time

Raw Scale

```{r}
data_long %>% 
  dplyr::select(subject, period, seizure.rate ) %>% 
  tidyr::spread(key = period,
                value = seizure.rate ) %>% 
  dplyr::select(-subject) %>% 
  cor() %>% 
  corrplot::corrplot.mixed()
```

Log Scale

```{r}
data_long %>% 
  dplyr::mutate(rate_wk = log(seizure.rate + 1)) %>% 
  dplyr::select(subject, period, rate_wk) %>% 
  tidyr::spread(key = period,
                value = rate_wk) %>% 
  dplyr::select(-subject) %>% 
  cor() %>% 
  corrplot::corrplot.mixed()
```



### Visualize

#### Oucome on the Raw Scale

There appear to be quite a few extreme values or outliers, particularly for the Progabide group during period one.

Since the outcome is truely a **COUNT**, we will model it with a **Poisson** distribution combined with a **LOG** link.

```{r}
data_long %>% 
  ggplot(aes(x = period,
             y = seizure.rate)) +
  geom_boxplot() +
  theme_bw() +
  facet_grid(.~ treatment)
```

To investigate possible outliers, we should transform the outcome with the `log` function first.  

> Note: Since some participants reported no seizures during a two week period and the `log(0)` is **unndefinded**, we must add some amount to the values before transforming.  Here we have chosen to add the value of $1$.

```{r}
data_long %>% 
  ggplot(aes(x = period,
             y = log(seizure.rate + 1))) +
  geom_boxplot() +
  
  theme_bw() +
  facet_grid(.~ treatment)
```

```{r}
data_long %>% 
  ggplot(aes(x = period,
             y = log(seizure.rate + 1))) +
  geom_line(aes(group = subject)) +
  theme_bw() +
  facet_grid(.~ treatment)
```


```{r}
data_long %>% 
  ggplot(aes(x = period,
             y = log(seizure.rate + 1))) +
  geom_smooth(aes(group = subject),
              method = "lm",
              se = FALSE) +
  geom_smooth(aes(group = 1),
              color = "red",
              size = 1.5,
              method = "lm",
              se = FALSE) +
  theme_bw() +
  facet_grid(.~ treatment)
```


## Poisson Regression (GLM)


> **Note:** THIS IS NEVER APPROPRIATE TO CONDUCT A GLM ON REPEATED MEASURES.  THIS IS DONE FOR ILLUSTRATION PURPOSES ONLY!
 
### Fit the model

```{r}
fit_glm <- glm(seizure.rate ~ base + age + treatment + offset(per),
               data = data_long,
               family = poisson(link = "log"))

summary(fit_glm)
```


### Tabulate the Model Parameters

```{r, results='asis'}
texreg::knitreg(list(fit_glm,
                     texreghelpr::extract_glm_exp(fit_glm,
                                                  include.any = FALSE)),
                custom.model.names = c("b (SE)", "IRR [95 CI]"),
                caption = "Poisson Generalized Linear Regression (GLM)",
                single.row = TRUE,
                digits = 3,
                ci.test = 1)
```


## Generalized Estimating Equations (GEE)


### Match Poisson Regresssion (GLM)

* correlation structure: `independence`
* scale parameter = $1$

```{r}
fit_gee_ind_s1 <- gee::gee(seizure.rate ~ base + age + treatment + offset(per),
                           data = data_long,
                           family = poisson(link = "log"),
                           id = subject,
                           corstr = "independence",
                           scale.fix = TRUE,
                           scale.value = 1)

summary(fit_gee_ind_s1)
```

* The estimates and the naive standard errors match the GLM exactly.  

* The naive SE's are much smaller (*half*) than the robust (sandwich) SE's, suggesting a poor fit.


### Change Correlation Sturucture

* correlation structure: `exchangeable`
* scale parameter = $1$

```{r}
fit_gee_exc_s1 <- gee::gee(seizure.rate ~ base + age + treatment + offset(per),
                           data = data_long,
                           family = poisson(link = "log"),
                           id = subject,
                           corstr = "exchangeable",
                           scale.fix = TRUE,
                           scale.value = 1)

summary(fit_gee_exc_s1)
```


* Although the estimated beta parameters are not much different, the naive SE's are some closer to the robust SE's.

#### Interpretation

> For determining significance, no p-values are given, however the p-value will be < .05 when the z-score is > 1.96.

"For this example, the estimates of standard errors under the independence are about half of the corresponding robust estimates, and the the situation improves only a little when an exchangeable structure is fitted.  Using naive standard errors leads, in particular, to a highly significant treatment effect which disappears when the robust estimates are used."




### Estimate the Additional Scale Parameter

"The problem with the GEE approach here, using either the independence or exchangeable correlation structure lies in constraining the scale parameter to be one.  For these data there is **overdispersion** *(variance is larger than the mean value)* which has to be accommodated by allowing this parameter to be freely estimated.  "


* correlation structure: `exchangeable`
* scale parameter = *freely estimated*



```{r}
fit_gee_exc_sf <- gee::gee(seizure.rate ~ base + age + treatment + offset(per),
                           data = data_long,
                           family = poisson(link = "log"),
                           id = subject,
                           corstr = "exchangeable",
                           scale.fix = FALSE)

summary(fit_gee_exc_sf)
```

* The naive SE's are much closer inline with the robust SE's.

* The scale parameter is estimated to be much larger than $1$.


"When this is done *(scale parameter is freely estimated)* it gives the last set of results shown above.  THe estimate of $\phi$ is 5.09 and the naive and robust estimates of the standard errors are now very similar.  "



### Compare Models


#### Raw Estimates (logit scale) 



```{r, results="asis"}
texreg::knitreg(list(fit_glm, 
                     fit_gee_ind_s1, 
                     fit_gee_exc_s1, 
                     fit_gee_exc_sf),
                custom.model.names = c("GLM",
                                       "GEE-Indep(1)",
                                       "GEE-Exchg(1)",
                                       "GEE-Exchg(free)"),
                caption = "Estimates on Logit Scale",
                digits = 3)
```



#### Exponentiate the Estimates (odds ratio scale)



```{r, results="asis"}
texreg::knitreg(list(extract_glm_exp(fit_glm), 
                     extract_gee_exp(fit_gee_ind_s1), 
                     extract_gee_exp(fit_gee_exc_s1), 
                     extract_gee_exp(fit_gee_exc_sf)),
                custom.model.names = c("GLM",
                                       "GEE-Indep(1)",
                                       "GEE-Exchg(1)",
                                       "GEE-Exchg(free)"),
                digits = 3,
                caption = "Estimates on Odds-Ratio Scale",
                caption.above = TRUE,
                ci.test = 1)
```

## Conculsion


**The Research Question**

The main question of interest is whether taking progabide reduced the number of epileptic seizures compared with placebo.

**The Conclusion**

There is no evidence that Progabide effects the weekly rate of epileptic seizures differently than placebo.










