# GLMM, Binary Outcome: Antibiotics for Leprosy

```{r, include=FALSE}
knitr::opts_chunk$set(comment     = "",
                      echo        = TRUE, 
                      warning     = FALSE, 
                      message     = FALSE,
                      fig.align   = "center", # center all figures
                      fig.width   = 6,        # set default figure width to 4 inches
                      fig.height  = 4)        # set default figure height to 3 inches
```

## Packages

### CRAN

```{r, message=FALSE, error=FALSE}
library(tidyverse)    # all things tidy
library(pander)       # nice looking genderal tabulations
library(furniture)    # nice table1() descriptives
library(texreg)       # Convert Regression Output to LaTeX or HTML Tables
library(psych)        # contains some useful functions, like headTail

library(lme4)         # Linear, generalized linear, & nonlinear mixed models
library(gee)          # Generalized Estimating Equations
library(effects)      # Plotting estimated marginal means

library(gridExtra)    # combining graphics
```


### GitHub

Helper `extract` functions for exponentiating parameters form generalized regression models within a `texreg` table of model parameters.

```{r, message=FALSE, error=FALSE}
# install.packages("devtools")
# library(devtools)
# install_github("SarBearSchwartz/texreghelpr")

library(texreghelpr)
```


## Data Prep

> Data on Obesity from the Muscatine Coronary Risk Factor Study.

**Source:** 

Table 10 (page 96) in Woolson and Clarke (1984). 
With permission of Blackwell Publishing.

**Reference:** 

Woolson, R.F. and Clarke, W.R. (1984). Analysis of categorical incompletel longitudinal data. Journal of the Royal Statistical Society, Series A, 147, 87-99.

**Description:**

The **Muscatine Coronary Risk Factor Study (MCRFS)** was a longitudinal study of coronary risk factors in school children in Muscatine, Iowa *(Woolson and Clarke 1984; Ekholm and Skinner 1998)*. Five cohorts of children were measured for `height` and `weight` in **1977**, **1979**, and **1981**. `Relative weight` was calculated as the **ratio** of a child's observed weight to the median weight for their age-sex-height group. Children with a relative weight greater than 110% of the median weight for their respective stratum were classified as `obese`. The analysis of this study involves binary data *(1 = obese, 0 = not obese)* collected at successive time points.


This data was also using in an article title *"Missing data methods in longitudinal studies: a review"* (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3016756/).


**Variable List:**

* Indicators   

    + `id` Child's unique identification number
    + `occas` Occasion number: 1, 2, 3
    
    
* Outcome or dependent variable    

    + `obesity` Obesity Status, 0 = no, 1 = yes  
    
    
* Main predictor or independent variable of interest   

    + `gender` 0 = Male, 1 = Female
    + `baseage` Baseline Age, mid-point of age-cohort 
    + `currage` Current Age, mid-point of age-cohort


### Import


```{r}
data_raw <- read.table("https://raw.githubusercontent.com/CEHS-research/data/master/MLM/Muscatine.txt", header=TRUE)

str(data_raw)

psych::headTail(data_raw, top = 10)
```



### Restrict to 350ID's of children with complete data for Class Demonstration

Dealing with missing-ness and its implications are beyond the score of this class.  Instead we are going to restrict our class analysis to a subset of 350 children who have complete data

> I am using the `set.seed()` function so that I can replicate the restults later.

```{r}
complete_ids <- data_raw %>% 
  dplyr::filter(obesity %in% c("0", "1")) %>%
  dplyr::group_by(id) %>% 
  dplyr::summarise(n = n()) %>% 
  dplyr::filter(n == 3) %>% 
  dplyr::pull(id)

set.seed(8892)

use_ids <- complete_ids %>% sample(350)

head(use_ids)
```


### Long Format


```{r}
data_long <- data_raw %>%  
  dplyr::filter(id %in% use_ids) %>% 
  mutate(id       = id     %>% factor) %>% 
  mutate(gender   = gender %>% factor(levels = 0:1, 
                                      labels = c("Male", "Female"))) %>% 
  mutate(age_base = baseage %>% factor) %>% 
  mutate(age_curr = currage %>% factor) %>% 
  mutate(occation = occas   %>% factor) %>% 
  mutate(obesity  = obesity %>% factor(levels = 0:1, 
                                       labels = c("No", "Yes"))) %>% 
  select(id, gender, age_base, age_curr, occation, obesity)

str(data_long)

psych::headTail(data_long, top = 10)
```


### Wide Format

```{r}
data_wide <- data_long %>% 
  gather(key = var,
         value = val,
         age_curr, obesity) %>% 
  unite(col = var_occ,
        var, occation) %>% 
  spread(key = var_occ,
         value = val) %>% 
  mutate_if(is.character, factor)%>% 
  group_by(id) %>% 
  mutate(num_miss = sum(is.na(c(obesity_1, obesity_2, obesity_3)))) %>% 
  ungroup() %>% 
  mutate(num_miss = as.factor(num_miss)) 


str(data_wide)

psych::headTail(data_wide, top = 10)
```













## Exploratory Data Analysis

### Summary Statistics

#### Demographics and Baseline

```{r}
data_wide %>% 
  furniture::table1(age_base, obesity_1,
                    splitby = ~ gender,
                    test    = TRUE,
                    na.rm   = FALSE,
                    output  = "html")
```


#### Status over Time


```{r}
data_summary <- data_long %>% 
  dplyr::group_by(gender, age_curr) %>% 
  dplyr::mutate(obesityN = case_when(obesity == "Yes" ~ 1,
                                     obesity == "No"  ~ 0)) %>% 
  dplyr::filter(complete.cases(gender, age_curr, obesityN)) %>% 
  dplyr::summarise(n = n(),
                   prob_est = mean(obesityN),
                   prob_SD  = sd(obesityN),
                   prob_SE  = prob_SD/sqrt(n))

data_summary
```



### Visualize

#### By cohort and gender

```{r}
data_long %>% 
  dplyr::group_by(gender, age_base, age_curr) %>% 
  dplyr::mutate(obesityN = case_when(obesity == "Yes" ~ 1,
                                     obesity == "No"  ~ 0)) %>% 
  dplyr::filter(complete.cases(gender, age_curr, obesityN)) %>% 
  dplyr::summarise(n = n(),
                   prob_est = mean(obesityN),
                   prob_SD  = sd(obesityN),
                   prob_SE  = prob_SD/sqrt(n)) %>% 
  ggplot(aes(x = age_curr,
             y = prob_est,
             group = age_base,
             color = age_base)) +
  geom_point() +
  geom_line() +
  theme_bw() + 
  labs(x = "Child's Age, years",
       y = "Proportion Obese") +
  facet_grid(. ~ gender)
```


#### BY only gender

```{r}
data_summary %>% 
  ggplot(aes(x = age_curr,
             y = prob_est,
             group = gender)) +
  geom_ribbon(aes(ymin = prob_est - prob_SE,
                  ymax = prob_est + prob_SE,
                  fill = gender),
              alpha = .3) +
  geom_point(aes(color = gender,
                 shape = gender)) +
  geom_line(aes(linetype = gender,
                color = gender)) +
  theme_bw() + 
  scale_color_manual(values = c("dodger blue", "hot pink")) +
  scale_fill_manual(values = c("dodger blue", "hot pink")) +
  labs(x = "Child's Age, years",
       y = "Proportion Obese")
```

Smooth out the trends

```{r}
data_summary %>% 
  ggplot(aes(x = age_curr,
             y = prob_est,
             group = gender,
             color = gender)) +
  geom_smooth(method = "lm",
              formula = y ~ poly(x, 2),
              se = FALSE) +
  theme_bw() + 
  scale_color_manual(values = c("dodger blue", "hot pink")) +
  scale_fill_manual(values = c("dodger blue", "hot pink")) +
  labs(x = "Child's Age, years",
       y = "Proportion Obese")
```



## Analysis Goal

> Does risk of obesity increase with age and are patterns of change similar for both sexes?

There are 5 age cohorts that were measured each for 3 years, baseage and currage are age midpoints of those cohort groups.  Which to include, current age or occasion?  **Assume no cohort effects.**  *If you do think this is an issue, include baseline age (`age_base`) and current age minus baseline age (`time`) in model.*


```{r}
data_long %>% 
  group_by(gender, age_base, occation) %>% 
  summarise(n       = n(),
            count   = sum(obesity == "Yes"),
            prop    = mean(obesity == "Yes"),
            se      = sd(obesity == "Yes")/sqrt(n)) %>%
  mutate(time = (occation %>% as.numeric) * 2 - 2) %>% 
  ggplot(aes(x = time,
             y = prop,
             color = gender,
             fill = gender))  +
  geom_ribbon(aes(ymin = prop - se,
                  ymax = prop + se),
              alpha = 0.2) +
  geom_point() +
  geom_line() +
  theme_bw() +
  facet_wrap(~ age_base, labeller = label_both) +
  labs(title    = "Observed Obesity Rates, by Gender within Cohort",
       subtitle = "Subset of 350 children with complete data",
       x        = "Time, years from 1977",
       y        = "Proportion of Children Characterized as Obese") +
  scale_fill_manual(values = c("dodgerblue3", "red")) +
  scale_color_manual(values = c("dodgerblue3", "red")) +
  scale_x_continuous(breaks = seq(from = 0, to = 4, by = 2)) +
  theme(legend.position = c(1, 0),
        legend.justification = c(1, 0),
        legend.background = element_rect(color = "black"))
```


```{r}
data_long %>% 
  group_by(gender, age_curr) %>% 
  summarise(n       = n(),
            count   = sum(obesity == "Yes"),
            prop    = mean(obesity == "Yes"),
            se      = sd(obesity == "Yes")/sqrt(n)) %>%
  ggplot(aes(x = age_curr %>% as.character %>% as.numeric,
             y = prop,
             group = gender,
             color = gender,
             fill = gender))  +
  geom_ribbon(aes(ymin = prop - se,
                  ymax = prop + se),
              alpha = 0.2) +
  geom_point() +
  geom_line() +
  theme_bw() +
  geom_vline(xintercept = 12, 
             linetype   = "dashed", 
             size       = 1, 
             color      = "navyblue") +
  labs(title    = "Observed Obesity Rates, by Gender (collapsing cohorts)",
       subtitle = "Subset of 350 children with complete data",
       x        = "Age of Child, years",
       y        = "Proportion of Children Characterized as Obese") +
  scale_fill_manual(values = c("dodgerblue3", "red")) +
  scale_color_manual(values = c("dodgerblue3", "red")) +
  scale_x_continuous(breaks = seq(from = 6, to = 18, by = 2)) +
  theme(legend.position = c(0, 1),
        legend.justification = c(-0.05, 1.05),
        legend.background = element_rect(color = "black"))
```


### Center time at twelve years old

```{r}
data_long <- data_long %>% 
  dplyr::mutate(age_center = age_curr %>% as.character %>% as.numeric -12) %>% 
  dplyr::mutate(obesity_num = obesity %>% as.numeric - 1)

psych::headTail(data_long)
```



## GLM Analysis

### Standard logistic regression

```{r}
fit_glm_1 <- glm(obesity_num ~ gender*age_center + gender*I(age_center^2), 
                 data   = data_long,
                 family = binomial)

fit_glm_2 <- glm(obesity_num ~ gender + age_center + I(age_center^2), 
                 data   = data_long,
                 family = binomial)
```


```{r}
texreg::screenreg(list(extract_glm_exp(fit_glm_1), 
                       extract_glm_exp(fit_glm_2)),
                  custom.model.names = c("Interaction",
                                         "Main Effects"),
                  caption = "GLM: Parameter EStimates",
                  single.row = TRUE,
                  ci.test = 1)
```

```{r}
plot_pred_glm <- Effect(c("gender", "age_center"), 
       fit_glm_2,
       xlevels = list(age_center = seq(from = -6, to = 6, by = 0.25))) %>% 
   data.frame %>%
   mutate(age = age_center + 12) %>% 
   ggplot(aes(x        = age,
              y        = fit,
              group    = gender,
              linetype = gender,
              fill     = gender,
              color    = gender)) +
  geom_ribbon(aes(ymin = fit - se,
                  ymax = fit + se),
              alpha = .3) +
   geom_line(size = 1.5) +
   theme_bw() +
   labs(title    = "Generalized Linear Model: Model #2",
        subtitle = "Predicted Probability of Obesity",
        x        = "Child's Age, years",
        y        = "Probability",
        linetype = "Gender",
        fill     = "Gender",
        color    = "Gender") +
   theme(legend.position = c(0, 1),
         legend.justification = c(-0.05, 1.05),
         legend.background = element_rect(color = "black"),
         legend.key.width = unit(2, "cm")) +
  scale_x_continuous(breaks = seq(from = 6, to = 18, by = 2))

plot_pred_glm 
```



## GEE Analysis

> ALWAYS: fix the scale parameter to 1 with binomial data!!!

```{r}
fit_gee_1in <- gee::gee(obesity_num ~ gender*age_center + gender*I(age_center^2), 
                        id          = id, 
                        data        = data_long,
                        family      = binomial,
                        corstr      = 'independence', 
                        scale.fix   = TRUE,
                        scale.value = 1)

fit_gee_1ex <- gee::gee(obesity_num ~ gender*age_center + gender*I(age_center^2), 
                        id          = id, 
                        data        = data_long,
                        family      = binomial,
                        corstr      = 'exchangeable', 
                        scale.fix   = TRUE,
                        scale.value = 1)

fit_gee_1un <- gee::gee(obesity_num ~ gender*age_center + gender*I(age_center^2), 
                        id          = id, 
                        data        = data_long,
                        family      = binomial,
                        corstr      = 'unstructured', 
                        scale.fix   = TRUE,
                        scale.value = 1)
```

```{r}
texreg::screenreg(list(extract_gee_exp(fit_gee_1in), 
                       extract_gee_exp(fit_gee_1ex),
                       extract_gee_exp(fit_gee_1un)),
                  custom.model.names = c("Independent",
                                         "Exchangable",
                                         "Unstructured"),
                  caption = "Gee Model Parameters: With Interactions",
                  single.row = TRUE,
                  ci.test = 1)
```

### Drop the interaction with `gender`.

```{r}
fit_gee_2in <- gee::gee(obesity_num ~ gender + age_center + I(age_center^2), 
                        id          = id, 
                        data        = data_long,
                        family      = binomial,
                        corstr      = 'independence', 
                        scale.fix   = TRUE,
                        scale.value = 1)

fit_gee_2ex <- gee::gee(obesity_num ~ gender + age_center + I(age_center^2), 
                        id          = id, 
                        data        = data_long,
                        family      = binomial,
                        corstr      = 'exchangeable', 
                        scale.fix   = TRUE,
                        scale.value = 1)

fit_gee_2un <- gee::gee(obesity_num ~ gender + age_center + I(age_center^2),  
                        id          = id, 
                        data        = data_long,
                        family      = binomial,
                        corstr      = 'unstructured', 
                        scale.fix   = TRUE,
                        scale.value = 1)
```


```{r}
texreg::screenreg(list(extract_gee_exp(fit_gee_2in), 
                       extract_gee_exp(fit_gee_2ex),
                       extract_gee_exp(fit_gee_2un)),
                  custom.model.names = c("Independent",
                                         "Exchangable",
                                         "Unstructured"),
                  caption = "Gee Model Parameters: Main Effects Only",
                  single.row = TRUE,
                  ci.test = 1)
```

### Select the **"final"** model.


```{r}
fit_geeglm_2un <- geepack::geeglm(obesity_num ~ gender + age_center + I(age_center^2),  
                                  id          = id, 
                                  data        = data_long,
                                  family      = binomial,
                                  corstr      = 'unstructured')

```




```{r}
plot_pred_gee <- expand.grid(gender = data_long$gender %>% levels,
                             age_center = seq(from = -6, to = 6, by = .1)) %>% 
   mutate(fit = predict(fit_geeglm_2un,
                        newdata = .,
                        type = "response")) %>%
  mutate(gender = fct_rev(gender)) %>% 
   mutate(age = age_center + 12) %>% 
   ggplot(aes(x        = age,
              y        = fit,
              group    = gender,
              linetype = gender,
              color    = gender)) +
   geom_line(size = 1.5) +
   theme_bw() +
   labs(title = "Generalized Estimating Equation: Model #2, unstructured",
        subtitle = "Predicted Probability of Obesity",
        x     = "Child's Age, years",
        y     = "Probability",
        linetype = "Gender",
        color = "Gender") +
   theme(legend.position = c(0, 1),
         legend.justification = c(-0.05, 1.05),
         legend.background = element_rect(color = "black"),
         legend.key.width = unit(2, "cm")) +
  scale_x_continuous(breaks = seq(from = 6, to = 18, by = 3))
  
  
plot_pred_gee
```



## GLMM Analysis

> IT IS GENERALLY NOT RECOMMENDED THAT RANDOM-SLOPES BE ESTIMATED FOR BINOMIAL GLMMS

```{r}
fit_glmm_1 <- lme4::glmer(obesity_num ~ age_center*gender + I(age_center^2)*gender + (1|id), 
                          data = data_long,
                          family = binomial) 

fit_glmm_2 <- lme4::glmer(obesity_num ~ gender + age_center + I(age_center^2) + (1|id), 
                          data = data_long,
                          family = binomial) 

```

Indicates smaller model is better, no interaction at higher level necessary

```{r}
anova(fit_glmm_1, fit_glmm_2)
```


```{r}
texreg::screenreg(list(extract_glmer_exp(fit_glmm_1), 
                       extract_glmer_exp(fit_glmm_2)),
                  custom.model.names = c("Interaction",
                                         "Main Effects"),
                  caption = "GLMM: Parameter EStimates",
                  single.row = TRUE,
                  ci.test = 1)
```




```{r}
Effect(c("gender", "age_center"),fit_glmm_2) %>% 
  data.frame %>% 
  mutate(fit_exp = exp(fit))
```




```{r}
plot_pred_glmm <- Effect(c("gender", "age_center"), 
       fit_glmm_2,
       xlevels = list(age_center = seq(from = -6, to = 6, by = 0.25))) %>% 
   data.frame %>%
   mutate(age = age_center + 12) %>% 
   ggplot(aes(x        = age,
              y        = fit,
              group    = gender,
              linetype = gender,
              color    = gender)) +
   geom_line(size = 1.5) +
   theme_bw() +
   labs(title    = "Generalized Linear Mixed Models: Model #2",
        subtitle = "Predicted Probability of Obesity",
        x        = "Child's Age, years",
        y        = "Probability",
        linetype = "Gender",
        color    = "Gender") +
   theme(legend.position = c(0, 1),
         legend.justification = c(-0.05, 1.05),
         legend.background = element_rect(color = "black"),
         legend.key.width = unit(2, "cm")) +
  scale_x_continuous(breaks = seq(from = 6, to = 18, by = 3))

plot_pred_glmm 
```

## Compare Methods

```{r}
texreg::screenreg(list(extract_glm_exp(fit_glm_2),
                       extract_gee_exp(fit_gee_2un),
                       extract_glmer_exp(fit_glmm_2)),
                  custom.model.names = c("GLM",
                                         "GEE",
                                         "GLMM"),
                  caption = "Compare Methods: Parameter EStimates",
                  single.row = TRUE,
                  ci.test = 1)
```


```{r, fig.height=10, fig.width=7.5}
gridExtra::grid.arrange(plot_pred_glm, 
                        plot_pred_gee, 
                        plot_pred_glmm)
```








