# GLMM, Count Outcome: Bolous



```{block type='rmdlink', echo=TRUE}
GLMM FAQ
by: Ben Bolker and others

https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html
```

```{r, include=FALSE}
knitr::opts_chunk$set(comment     = "",
                      echo        = TRUE, 
                      warning     = FALSE, 
                      message     = FALSE,
                      fig.align   = "center", # center all figures
                      fig.width   = 6,        # set default figure width to 4 inches
                      fig.height  = 4)        # set default figure height to 3 inches
```

## Packages

### CRAN

```{r, message=FALSE, error=FALSE}
library(tidyverse)    # all things tidy
library(pander)       # nice looking genderal tabulations
library(furniture)    # nice table1() descriptives
library(texreg)       # Convert Regression Output to LaTeX or HTML Tables
library(psych)        # contains some useful functions, like headTail

library(lme4)         # Linear, generalized linear, & nonlinear mixed models
library(gee)          # Generalized Estimating Equations
library(optimx)       # Optimizers for use in lme4::glmer
library(MuMIn)        # Multi-Model Inference (caluclate QIC)
library(effects)      # Plotting estimated marginal means

library(gridExtra)    # combining graphics
```


### GitHub

Helper `extract` functions for exponentiating parameters form generalized regression models within a `texreg` table of model parameters.

```{r, message=FALSE, error=FALSE}
# install.packages("devtools")
# library(devtools)
# install_github("SarBearSchwartz/texreghelpr")

library(texreghelpr)
```


## Data Prep

> Bolus data from Weiss 2005

Patient controlled analgesia (PCA) comparing 2 dosing regimes to self-control pain 


The dataset has the number of requests per interval in 12 successive four-hourly intervals following abdominal surgery for 65 patients in a clinical trial to compare two groups (bolus/lock-out combinations).

Bolus=  Large dose of medication given (usually intravenously by direct infusion injection or gravity drip) to raise blood-level concentrations to a therapeutic level

A ‘lock-out’ period followed each dose, where subject may not administer more medication.  

* Lock-out time is twice as long in 2mg/dose group
* Allows for max of 30 dosages in 2mg/dose and 60 in 1mg/dose group in any 4-hour period
* No responses neared these upper limits



**Variable List:**

* Indicators   

    + `id` unique subject indicator
    + `time` 11 consecutive 4-hour periods: 0, 1, 2, ..., 10


* Outcome or dependent variable    

    + `count` # doses recorded for in that 4-hour period


* Main predictor or independent variable of interest   

    + `group`1mg/dose group, 2mg/dose group



### Import

```{r}
data_raw <- read.table("https://raw.githubusercontent.com/CEHS-research/data/master/MLM/bolus.txt",
  header = TRUE
)

str(data_raw) # see the structure

psych::headTail(data_raw)
```

### Long Format

```{r}
data_long <- data_raw %>% 
  dplyr::mutate(id = factor(id)) %>% 
  dplyr::mutate(timeF = factor(time))

str(data_long) # see the structure

psych::headTail(data_long)
```

### Wide Format

```{r}
data_wide <- data_long %>% 
  dplyr::select(-timeF) %>% 
  tidyr::spread(key = time,
                value = count,
                sep = "_")

str(data_wide) # see the structure

psych::headTail(data_wide)
```


## Exploratory Data Analysis

### Summary Statistics

#### Demographics and Baseline

```{r}
data_wide %>% 
  furniture::table1(time_0,
                    splitby = ~ group,
                    test    = TRUE,
                    na.rm   = FALSE,
                    output  = "html")
```


#### Status over Time


```{r}
data_wide %>% 
  furniture::table1(time_0, 
                    time_1, time_2, time_3, time_4, time_5, 
                    time_6, time_7, time_8, time_9, time_10,
                    splitby = ~ group,
                    test    = TRUE,
                    na.rm   = FALSE,
                    output  = "html")
```


#### Correlation over Time

```{r}
data_wide %>% 
  dplyr::select(starts_with("time_")) %>% 
  cor() %>% 
  corrplot::corrplot.mixed()
```



### Visualize

#### Person Profile Plots, by Group

```{r}
data_long %>% 
  ggplot(aes(x = time, 
             y = count)) +
  geom_line(aes(group = id),
              color = "gray") +
  facet_grid(. ~ group) + 
  theme_bw()
```



```{r}
data_long %>% 
  ggplot(aes(x = time, 
             y = count)) +
  geom_smooth(aes(group = id),
              method = "lm",
              color = "gray",
              se = FALSE) +
  facet_grid(. ~ group) + 
  theme_bw()
```



```{r}
data_long %>% 
  ggplot(aes(x = time, 
             y = count)) +
  geom_smooth(aes(group = id),
              method = "lm",
              formula = y ~ poly(x, 2),
              color = "gray",
              se = FALSE) +
  facet_grid(. ~ group) + 
  theme_bw()
```




```{r}
data_long %>% 
  ggplot(aes(x = time, 
             y = count)) +
  geom_line(aes(group = id)) +
  geom_smooth(method = "loess",
              color = "blue",
              se = FALSE) +
  geom_smooth(method = "lm",
              color = "red",
              se = FALSE) +
  facet_grid(. ~ group) + 
  theme_bw()
```

```{r}
data_long %>% 
  ggplot(aes(x = time, 
             y = log(count + .1))) +
  geom_line(aes(group = id)) +
  geom_smooth(method = "loess",
              color = "blue",
              se = FALSE) +
  geom_smooth(method = "lm",
              color = "red",
              se = FALSE) +
  facet_grid(. ~ group) + 
  theme_bw()
```


```{r}
data_long %>% 
  ggplot(aes(x = time, 
             y = count)) +
  geom_smooth(method = "loess",
              color = "gray",
              se = FALSE) +
  geom_smooth(method = "lm",
              color = "red",
              se = FALSE) +
  geom_smooth(method = "lm",
              formula = y ~ poly(x, 2),
              color = "green",
              se = FALSE) +
  geom_smooth(method = "lm",
              formula = y ~ poly(x, 3),
              color = "purple",
              se = FALSE) +
  facet_grid(. ~ group) + 
  theme_bw()
```

#### Means over Time, BY Group

```{r}
data_long %>% 
  dplyr::group_by(group, timeF) %>% 
  dplyr::summarise(M = mean(count)) %>% 
  ggplot(aes(x = timeF,
             y = M,
             group = group,
             color = group)) +
  geom_point() + 
  geom_line() + 
  theme_bw() + 
  labs(x = "Four-week Intervals",
       y = "Sample Mean Count",
       color = "Dose") 
```


```{r}
data_long %>% 
  dplyr::group_by(group, timeF) %>% 
  dplyr::summarise(M = mean(log(count + .1))) %>% 
  ggplot(aes(x = timeF,
             y = M,
             group = group,
             color = group)) +
  geom_point() + 
  geom_line() + 
  theme_bw() + 
  labs(x = "Four-week Intervals",
       y = "Sample Mean Count",
       color = "Dose") 
```


## GEE Analysis

### Fit Various Correlation Structures

#### Exchangable

```{r}
mod_gee_ex <- gee::gee(count ~ group + time + I(time^2),
                       id = id,
                       data = data_long,
                       family = poisson(link = "log"),
                       corstr = 'exchangeable')

summary(mod_gee_ex)
```

#### Auto-regressive (correlation decay over time)

```{r}
mod_gee_ar <- gee::gee(count ~ group + time + I(time^2),
                       id = id,
                       data = data_long,
                       family = poisson(link = "log"),
                       corstr = "AR-M",
                       Mv = 1)	

summary(mod_gee_ar)
```

#### Unstructured

```{r}
mod_gee_un <- gee::gee(count ~ group + time + I(time^2),
                       id = id,
                       data = data_long,
                       family = poisson(link = "log"),
                       corstr = "unstructured")	

summary(mod_gee_un)
```

### Compare Corelation Structures

#### QIC: Model Fit 

```{r}
MuMIn::model.sel(mod_gee_ex, 
                 mod_gee_ar, 
                 mod_gee_un, 
                 rank = "QIC")    #sorts the best to the TOP, uses QIC(I) to choose corelation structure
```


#### Model Parameter Table

```{r, include = FALSE}
texreg::screenreg(list(extract_gee_exp(mod_gee_ex),
                       extract_gee_exp(mod_gee_ar),
                       extract_gee_exp(mod_gee_un)),
                  custom.model.names = c("Exchangable",
                                         "AutoRegressive",
                                         "Unstructured"),
                  caption = "GEE: Main Effects Only, with Quadratic Time",
                  single.row = TRUE,
                  ci.test = 1)
```


```{r, results='asis'}
# Knit to Website: texreg::htmlreg()
# Knit to PDF:     texreg::texreg()
# View on Screen:  texreg::screenreg()

texreg::htmlreg(list(extract_gee_exp(mod_gee_ex),
                     extract_gee_exp(mod_gee_ar),
                     extract_gee_exp(mod_gee_un)),
                custom.model.names = c("Exchangable",
                                       "AutoRegressive",
                                       "Unstructured"),
                caption = "GEE: Main Effects Only, with Quadratic Time",
                single.row = TRUE,
                ci.test = 1)
```

### Add Interaction Terms

```{r}
mod_gee_ar2 <- gee::gee(count ~ group*time + group*I(time^2),
                        id = id,
                        data = data_long,
                        family = poisson(link = "log"),
                        corstr = "AR-M",
                        Mv = 1)	

summary(mod_gee_ar2)
```


```{r}
MuMIn::QIC(mod_gee_ar, 
           mod_gee_ar2, 
           typeR = TRUE)    # NOT the default
```

```{r}
texreg::screenreg(list(extract_gee_exp(mod_gee_ar),
                       extract_gee_exp(mod_gee_ar2)),
                  single.row = TRUE,
                  ci.test = 1)
```




### Visualize Best Model

#### Model Parameter Table

```{r, include=FALSE}
texreg::screenreg(extract_gee_exp(mod_gee_ar),
                  single.row = TRUE,
                  ci.test = 1)
```


```{r, results='asis'}
# Knit to Website: texreg::htmlreg()
# Knit to PDF:     texreg::texreg()
# View on Screen:  texreg::screenreg()

texreg::htmlreg(extract_gee_exp(mod_gee_ar),
                caption = "GEE: Final Model (auto regressive)",
                single.row = TRUE,
                ci.test = 1)
```

#### Refit via `geepack::geeglm()`

```{r}
mod_geeglm_ar <- geepack::geeglm(count ~ group + time + I(time^2),
                       id = id,
                       data = data_long,
                       family = poisson(link = "log"),
                       corstr = "ar1")	

summary(mod_geeglm_ar)
```


#### Predict over a Grid 

Estimated Marginal Mean Counts

```{r}
expand.grid(time = 0:10,
            group = levels(data_long$group)) %>% 
  dplyr::mutate(fit = predict(mod_geeglm_ar,
                              newdata = .,
                              type = "response")) %>% 
  tidyr::spread(key = group,
                value = fit)
```

#### Plot Estimated Marginal Means

```{r}
expand.grid(time = 0:10,
            group = levels(data_long$group)) %>% 
  dplyr::mutate(fit = predict(mod_geeglm_ar,
                              newdata = .,
                              type = "response")) %>% 
  ggplot(aes(x = time,
             y = fit,
             color = fct_rev(group))) +
  geom_point() + 
  geom_line() + 
  theme_bw() + 
  labs(x = "Four-week Intervals",
       y = "Estimated Populaton Mean Count",
       color = "Dose") + 
  scale_x_continuous(breaks = 0:10)
```



## GLMM Analysis


### RI: Random Intercepts Only

```{r}
mod_glmer_ri <- lme4::glmer(count ~ group + I(time/4) + I((time/4)^2) + (1|id),
                            data = data_long,
                            family = poisson(link = "log"))

summary(mod_glmer_ri)
```


### RIAS: Random Intercepts and Slopes

```{r}
mod_glmer_rias <- lme4::glmer(count ~ group + I(time/4) + I((time/4)^2) + (I(time/4)|id),
                              data = data_long,
                              family = poisson(link = "log"))

summary(mod_glmer_rias)
```

 

```{r}
anova(mod_glmer_ri, mod_glmer_rias)
```

```{r, include=FALSE}
# Knit to Website: texreg::htmlreg()
# Knit to PDF:     texreg::texreg()
# View on Screen:  texreg::screenreg()

texreg::screenreg(list(extract_glmer_exp(mod_glmer_ri),
                       extract_glmer_exp(mod_glmer_rias)),
                  custom.model.names = c("Intecepts", "Intercepts and Slopes"),
                  caption = "GLMM: Compare Random Effects",
                  single.row = TRUE,
                  ci.test = 1)
```

```{r, results="asis"}
# Knit to Website: texreg::htmlreg()
# Knit to PDF:     texreg::texreg()
# View on Screen:  texreg::screenreg()

texreg::htmlreg(list(extract_glmer_exp(mod_glmer_ri),
                     extract_glmer_exp(mod_glmer_rias)),
                custom.model.names = c("Intecepts", "Intercepts and Slopes"),
                caption = "GLMM: Compare Random Effects",
                single.row = TRUE,
                ci.test = 1)
```



### RAIS: Add Interaction


See the [GLMM - Optimizers page](https://cehs-research.github.io/eBook_multilevel/glmm-binary-outcome-antibiotics-for-leprosy.html#glmm---optimizers) for more information on convergence problems.

```{r}
mod_glmer_rias2 <- lme4::glmer(count ~ group*I(time/4) + group*I((time/4)^2) + ( I(time/4)|id ),
                              data = data_long,
                              control = glmerControl(optimizer ="Nelder_Mead"),  # convergence issues resolved
                              family = poisson(link = "log"))	
```

```{r}
anova(mod_glmer_rias, mod_glmer_rias2)
```



### Visualize Best Model

#### Model Parameter Table

```{r, include=FALSE}
texreg::screenreg(list(extract_glmer_exp(mod_glmer_rias)),
                  single.row = TRUE,
                  ci.test = 1)
```


```{r, results='asis'}
# Knit to Website: texreg::htmlreg()
# Knit to PDF:     texreg::texreg()
# View on Screen:  texreg::screenreg()

texreg::htmlreg(list(extract_glmer_exp(mod_glmer_rias)),
                caption = "GLMM: Final Model",
                single.row = TRUE,
                ci.test = 1)
```



#### Estimated Marginal Means

```{r}
effects::Effect(focal.predictors = c("group", "time"),
                xlevels = list(time = 0:10),
                mod = mod_glmer_rias) %>% 
  data.frame %>% 
  dplyr::select(group, time, fit) %>% 
  tidyr::spread(key = group,
                value = fit,
                sep = "_")
```

#### Plot Estimated Marginal Means

```{r}
effects::Effect(focal.predictors = c("group", "time"),
                xlevels = list(time = seq(from = 0, to = 10, by = .25)),
                mod = mod_glmer_rias) %>% 
  data.frame %>% 
  ggplot(aes(x = time,
             y = fit)) +
  geom_ribbon(aes(ymin = fit - se,
                  ymax = fit + se,
                  fill = group),
              alpha = .3) +
  geom_line(aes(color = group)) + 
  theme_bw() + 
  labs(x = "Four-week Intervals",
       y = "Estimated Populaton Mean Count",
       color = "Dose",
       fill = "Dose") + 
  scale_x_continuous(breaks = 0:10)
```


```{r}
data_long %>% 
  dplyr::mutate(fit = predict(mod_glmer_rias,
                              newdata = .,
                              type = "response")) %>% 
  ggplot(aes(x = time, 
             y = fit)) +
  geom_line(aes(group = id)) +
  facet_grid(.~ group) +
  theme_bw()
```

